---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff - changed by Alba"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

# 1. Working with the dplyr and ggplot2 packages

### 0.1 How to filter outliers and use ggplot

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# restore your library (keep at the start of every script)
renv::restore()
# clear the R environment
rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

You can also use the function read.csv() it will print the first 10 rows
of your data set with all rows.

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")
print(FactCockleSize) # to show contents of data
names(FactCockleSize) # names of variables
nrow(FactCockleSize) # number of rows
```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}
ggplot2::ggplot(data=FactCockleSize, # what data you're using
                mapping=aes(x=length_mm,y=thickness_mm)) + # what is on the axis
  geom_point()
# there is a big outlier, likely a data entry mistake, remove it
  FactCockleSize |> dplyr::filter(thickness_mm>500)
  
  FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> # means pipe
    dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469) |> # filter = select particular rows. When ID of cockle outliers identified, remove them
    dplyr::mutate(year=factor(year)) # make a year factor
# plot the data
  ggplot2::ggplot(data=FactCockleSize,
               mapping=aes(x=length_mm,y=thickness_mm)) + geom_point()
# find that other outlier
  FactCockleSize |> dplyr::filter(thickness_mm>10 & length_mm<5)
# also remove 469 above
```

dplyr:

-   filter: select particular rows

-   select: select particular variable columns

-   mutate: transform and calculate variable

-   group_by: make group

-   summarize: creates a new data frame

Further explore the plot with a regression line through all the data

```{r}
# plot the data with a regression line
ggplot2::ggplot(data=FactCockleSize,
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point() + 
  geom_smooth(method="lm") # added regreesion line
# calculate a linear regression
model_lm<-lm(thickness_mm~length_mm, data=FactCockleSize) # line regression where y is explained by x -> correspond it to variables then add from what data it is from with data=
summary(model_lm)
# equation: thickness_mm=-0.353+0.669*lenth_mm
# a cockle of length 20mm has a predicted width of:
-0.353+0.669*20
# this regression is significant, R2=0.90, P<0.001 (***)
```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line
ggplot2::ggplot(data=FactCockleSize,
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point(mapping=aes(col=year)) + # color only applies to this, make color a factor
  geom_smooth(method="lm")
```

Make a panel plot where with each year is shown as a separate graph

```{r}
#first method
drop_na(FactCockleSize) |> #remove missing data
  ggplot2::ggplot(mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point(mapping=aes(col=year)) + 
  geom_smooth(method="lm") + 
  facet_wrap(~year)

drop_na(FactCockleSize) |>
ggplot2::ggplot(mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point(mapping=aes(col=year), size=2) + 
  geom_smooth(method="lm") + 
  xlab("cockle length (mm)") + #lab = labels
  ylab("cockle width (mm)") + 
  facet_wrap(~year)
```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is abou the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
